---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
    keep_md: true
    author: "Vannia Gonzalez"
date: "November 21, 2016"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='figure/plot-', cache=TRUE)
```
## Loading and preprocessing the data

```{r,results='hide'}
data<-read.csv("activity.csv",sep=",",header = TRUE,stringsAsFactors = FALSE)
library(lubridate)
data$date<-as.POSIXct(data$date,format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Total steps per day
```{r}
daily_sum<-aggregate(data$steps ~ strftime(data$date, "%m-%d"), FUN = sum,na.rm=TRUE)
#steps_daily<-data%>%group_by(date)%>%summarise(tot=sum(steps,na.rm=TRUE))   //uses dplyr,plyr
```

Histogram: steps per day
```{r,1,fig.cap="Histogram 1"}
hist(daily_sum$`data$steps`,xlab="Total number of steps each day",main="Histogram of total number of steps per day")
abline(v=mean(daily_sum$`data$steps`),col="blue")
abline(v=median(daily_sum$`data$steps`),col="red")
```
The blue and red vertical lines show the values for the mean and median, respectively, for the average steps each day, given explicitely by:  

**1. Mean**
```{r}
mean(daily_sum$`data$steps`,na.rm=TRUE)
```
**2. Median**
```{r}
median(daily_sum$`data$steps`,na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r}
avg_interval<-setNames(aggregate(data$steps~data$interval,FUN=mean,na.rm=TRUE),c("int","avg"))
```
```{r,2}
with(avg_interval,plot(int,avg,type="l",xlab="time interval",ylab="Average steps",main="Average steps per interval",col="blue"))
m<-avg_interval$int[which(avg_interval$avg==max(avg_interval$avg))]
```
  
**Interval with maximum number of average steps across all days: `r m`**


## Imputing missing values

```{r}
m<-sum(is.na(data$steps))
```
Total number of missing values: `r m`


In the next step I fill the missing values in the steps variable with the average steps (across all days) made within the corresponding interval 
```{r}
data2<-data
for(i in 1:nrow(data2)){ 
        if(is.na(data2$steps[i])){
                inum<-data2[i,3] #nterval number
                data2$steps[i]<-avg_interval[avg_interval$int==inum,2]
        }
        
}
```
Verify that the total missing values are now zero.
```{r}
sum(is.na(data2$steps))
```

Histogram with new dataset

```{r,3,fig.cap="Histogram 2"}
daily_sum2<-aggregate(data2$steps ~ strftime(data2$date, "%m-%d"), FUN = sum,na.rm=TRUE)
hist(daily_sum2$`data2$steps`,xlab="Total number of steps each day",main="Histogram of total number of steps per day (replacing missing values)")
abline(v=mean(daily_sum2$`data2$steps`),col="blue")
abline(v=median(daily_sum2$`data2$steps`),col="red")
abline(v=mean(daily_sum$`data$steps`),col="orange")
abline(v=median(daily_sum$`data$steps`),col="green")
```
The former plot shows a histogram of the average steps per day when replacing the missing values for the step variable with a reasonable value such as the average steps across all days for the corresponding interval for each missing value (dataset 2). Note that this histogram is not significantly different from that, ignoring the missing values (first figure of this report).
I plotted in different colors the means and medians for both datasets but these are apparently indistinguishable (they overlap).

Explicitly the mean and median are given by:

1. Mean for second dataset
```{r}
mean(daily_sum2$`data2$steps`)
```

2. Median for second dataset
```{r}
median(daily_sum2$`data2$steps`)
```

If we compare both datasets quantiles we see there is no significant impact in the set of values for each interval limit.

```{r}
summary(daily_sum$`data$steps`)
summary(daily_sum2$`data2$steps`)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
avg_interval<-setNames(aggregate(data2$steps~data2$interval+data2$date,FUN=mean),c("int","wtype","avg"))
```

```{r,4}
library(lattice)
xyplot(avg~int|wtype,data=avg_interval, 
        type="l", 
        ylab="Time interval", xlab="Average steps",
	      layout=c(1,2) 
)
```
